PANCHANGE_data_preprocessing

### 2020-07-21 PANCHANGE_data_preprocessing2020-07-15.Rmd
Transfered all comments in script into action points on github

This will be the master script to run all analyses for the project "Anxiety, depression and trauma symptom change during the COVID-19 pandemic: retrospective versus objective assessment" - Young et al (2020)

@@

# Colour palettes
Define colours for plotting this are the standard coping colours
+++CH: Is there a document in which these colours are defined?
```{r Colour palettes: COPING}
COPINGpalette2 <- c("#78D9C5",
                    "#F5BE5E")

@@

# Data import and merging
+++CH04-07-20: I have added all read-ins of the data. How will the coping data be stored? -> Question for Yuhao
+++CH04-07-20: We need to discuss the naming of the variables PHQ9 vs phq
+++CH04-07-20: We need to discuss the naming of the variables .0 vs .retro

GLAD only receives current measures and RAMP questionnaire

All COPING surveys are separate surveys
+++TOPHER
***
++KT: Need to make sure that the startDate's for each survey are coded accordingly, e.g. startDate.glad, startDate.coping, startDate.RAMP. Otherwise people who are in both GLAD and COPING will have two dates.
+++CH: I have added this

@@

Clean Age variable
Age outlier
```{r age outlier}
# define age limits 
age_upper_limit = 117 # oldest person in the world is 117 years
age_lower_limit = 16

@@

```{r glad dem head}
head(glad.dem.raw.id)
```
++KT: The age variable here needs to be cleaned - have added a chunk above (before age grouping) for this. Before: min = -55 and max = 290. 
```{r glad dem summary}
summarytools::dfSummary(
  glad.dem.raw.id,
  graph.col = F,
  valid.col = F,
  labels.col = F)
```

@@

# Grouping mental health disorders
*#Disorder groupings - from COPING/RAMP analysis minutes*
1. Depressive disorder group: major depression, antenatal/postnatal depression, premenstrual depression.
2. Anxiety group: generalised anxiety, social anxiety, specific phobia, agoraphobia, panic_disorder, panic_attacks. 
3. Agoraphobia and panic_disorder group: panic_disorder, agoraphobia, panic_attacks.
4. Eating disorders group: anorexianervosa, atypical anorexia nervosa, bulimia nervosa, overeating or bingeeating.
5. Obsessive complusive disorders: obsessive compulaive disorder, obsessive compulsive related disorders. 
6. Psychotic disorders: schizophrenia, schizoaffective disorder, general psychosis or psychotic illness. 
7. Personality disorders: cluster A, cluster B, cluster C
8. Control group: no diagnosed disorders.

@@

Depressive disorders group
+++CH: NEEDS BETTER LOGIC FOR MALES!!!!
++KT: Sex issue fixed - code has been checked. 407 people with no MDD but have either postnatal/pmdd.
```{r glad mhq depressive_disorders}
glad.mhq.raw.id <- glad.mhq.raw.id %>%
  mutate(
    depressive_disorders_numeric =
      case_when(
        (mhq.depression_numeric == "0" &
        (mhq.pregnancy_depression_numeric == "0" | is.na(mhq.pregnancy_depression_numeric)) & 
        (mhq.premenstrual_dysphoric_disorder_pmdd_numeric == "0" | is.na(mhq.premenstrual_dysphoric_disorder_pmdd_numeric))) ~ 0,
        mhq.depression_numeric == "1" ~ 1,
        mhq.pregnancy_depression_numeric == "1" ~ 1,
        mhq.premenstrual_dysphoric_disorder_pmdd_numeric == "1" ~ 1))

@@

```{r Check agoraphobia_panic_disorder}
glad.mhq.raw.id %>%
  select(
    agoraphobia_panic_disorder,
    agoraphobia_panic_disorder_numeric,
    mhq.agoraphobia_numeric,
    mhq.panic_disorder_numeric,
    mhq.panic_attacks_numeric,
    mhq.agoraphobia,
    mhq.panic_disorder,
    mhq.panic_attacks
  )
```
+++CH: Why are there so many NAs in mhq.panic_disorder?
++KT: Added late to questionnaire.

@@

+++CH: Lots of NAs in schizoaffective_disorder 
```{r Check psychotic_disorders}
glad.mhq.raw.id %>%
  select(
    psychotic_disorders,
    psychotic_disorders_numeric,
    mhq.schizophrenia_numeric,
    mhq.schizoaffective_disorder_numeric,
    mhq.type_psychosis_psychotic_illness_numeric,
    mhq.schizophrenia,
    mhq.schizoaffective_disorder,
    mhq.type_psychosis_psychotic_illness
  )
```

@@

Create new dataframe with the derived categories

@@

+++ ADD 3 CHUNKs for phq, gad for each comparison
1) prepandemic - retro -> most interesting (in analysis needs to be adjusted for time difference between prepandemic and coping sign up)
2) prepandemic - base
3) retro - base

+++ ADD a case_when CHUNKs - who has what data - to get accurate Ns. 
1) binary coded has only prepandemic == 0
2) prepandemic and baseline == 1
3) prepandemic, baseline, and retrospective == 2

+++ chunks for missingness
1) per individual
2) per questionnaire item/total questionnaire

@@

Retro for the PCL does not mean retrospective
```{r coping glad pcl cleaning}
coping.glad.pcl.raw.id <- coping.glad.pcl.raw %>%
  drop_na(externalDataReference) %>% # Drop NAs
  distinct(externalDataReference, .keep_all = TRUE) %>% # Remove duplicates based on ID
  separate(externalDataReference, into = c("Sample", "ID"), sep = 4) %>% # Split ID in Sample and Number
  mutate(ID = as.numeric(ID),
         startDate.glad = startDate) %>% # Rename startDate
  select(
    Sample, # Sample
    ID, # ID
    pcl.stressful_experience_repeated_images_base_numeric = pcl6.stressful_experience_repeated_images_numeric,
    pcl.stressful_experience_repeated_images_base = pcl6.stressful_experience_repeated_images,
    pcl.stressful_experience_upset_reminded_base_numeric = pcl6.stressful_experience_upset_reminded_numeric,
    pcl.stressful_experience_upset_reminded_base = pcl6.stressful_experience_upset_reminded,
    pcl.stressful_situation_avoiding_activities_base_numeric = pcl6.avoiding_activities_stressful_experience_numeric, # differ to GLAD prepan
    pcl.stressful_situation_avoiding_activities_base = pcl6.avoiding_activities_stressful_experience, # differ to GLAD prepan
    pcl.cut_people_feeling_distant_base_numeric = pcl6.cut_people_feeling_distant_numeric,
    pcl.cut_people_feeling_distant_base = pcl6.cut_people_feeling_distant,
    pcl.feeling_irritable_or_having_angry_outbursts_base_numeric = pcl6.feeling_irritable_or_having_angry_outbursts_numeric,
    pcl.feeling_irritable_or_having_angry_outbursts_base = pcl6.feeling_irritable_or_having_angry_outbursts,
    pcl.difficulty_concentrating_base_numeric = pcl6.difficulty_concentrating_numeric,
    pcl.difficulty_concentratin_base = pcl6.difficulty_concentrating,
    pcl.stressful_experience_repeated_images_retro_numeric = pcl6.stressful_experience_repeated_images.1_numeric,
    pcl.stressful_experience_repeated_images_retro = pcl6.stressful_experience_repeated_images.1,
    pcl.stressful_experience_upset_reminded_retro_numeric = pcl6.stressful_experience_upset_reminded.1_numeric,
    pcl.stressful_experience_upset_reminded_retro = pcl6.stressful_experience_upset_reminded.1,
    pcl.stressful_situation_avoiding_activities_retro_numeric = pcl6.avoiding_activities_stressful_experience.1_numeric,
    pcl.stressful_situation_avoiding_activities_retro = pcl6.avoiding_activities_stressful_experience.1,
    pcl.cut_people_feeling_distant_retro_numeric = pcl6.cut_people_feeling_distant.1_numeric,
    pcl.cut_people_feeling_distant_retro = pcl6.cut_people_feeling_distant.1,
    pcl.feeling_irritable_or_having_angry_outbursts_retro_numeric = pcl6.feeling_irritable_or_having_angry_outbursts.1_numeric,
    pcl.feeling_irritable_or_having_angry_outbursts_retro = pcl6.feeling_irritable_or_having_angry_outbursts.1,
    pcl.difficulty_concentrating_retro_numeric = pcl6.difficulty_concentrating.1_numeric,
    pcl.difficulty_concentratin_retro = pcl6.difficulty_concentrating.1
  ) %>%
  mutate_if(is.numeric, ~na_if(., -88)) %>% # Recode missing values to NAs in numeric variables
  mutate_if(is.numeric, ~na_if(., -99)) %>%
  mutate_if(is.numeric, ~na_if(., -77)) %>%
  mutate_if(is.factor, ~na_if(., "Seen but not answered")) %>% # Recode missing values to NAs in factor variables
  mutate_if(is.factor, ~na_if(., "Don't know")) %>%
  mutate_if(is.factor, ~na_if(., "Prefer not to say")) %>%
  mutate_if(is.factor, ~na_if(., "Prefer not to answer")) %>%
  mutate_if(is.factor, ~forcats::fct_drop(., only = "Seen but not answered")) %>% # Drop empty factor levels
  mutate_if(is.factor, ~forcats::fct_drop(., only = "Don't know")) %>%
  mutate_if(is.factor, ~forcats::fct_drop(., only = "Prefer not to answer")) %>%
  mutate_if(is.factor, ~forcats::fct_drop(., only = "Prefer not to say"))

@@

```{r coping glad phq summary labels}
summarytools::dfSummary(
  coping.glad.phq.raw,
  graph.col = F,
  valid.col = F,
  #  labels.col = F
)
```

@@

Exclusion of inflammatory bowel disease (IBD) BioResource (BR) cohort
+++CH: or alternatively sample according to IBD prevalence in UK population
+++CH: What is the IBD prevalence in the UK population? -> Literature search
+++CH: This chunk needs to be included when we have the data available from the NBR
```{r COVID exclusion of IBD}
#dim(dat.no.sex.no.age)
#
#dat.no.sex.no.age <- dat.no.sex.no.age %>%
#  filter(NBR.IBD == 0)
#
#dim(dat.no.sex.no.age.no.ibd)
#
#dim(dat.no.sex.no.age)[1]-dim(dat.no.sex.no.age.no.ibd)[1]

@@

#Missigness of questionnaire items
+++CH: We need to discuss how we want to exclude these participants that are missing answers to the questionnaires
GAD7 ?
PHQ9 ? 
PCL-6?
+++CH: Check for literature
```{r Exclusion of participants because of missing questionnaire items}
#dim(dat.no.sex.no.age.no.ibd)
#dat.no.sex.no.age.no.ibd.incomplete.questionnaire <- dat.no.sex.no.age.no.ibd %>%
#  filter(#ADD CONDITION HERE)
#dim(dat.no.sex.no.age.no.ibd.incomplete.questionnaire)
#
#dim(dat.no.sex.no.age.no.ibd)[1]-dim(dat.no.sex.no.age.no.ibd.incomplete.questionnaire)[1]

@@

# Definition of prepandemic and baseline 
Exlcude people who have joined GLAD/EDGI after the pandemic began (i.e., cut off that we choose)
Potentially GLAD/EDGI particiapnts between January to April. EDGI launched on February 24th.

These participants with sign ups later than these dates could  potentially be counted as COPING baseline.

@@

Retrospective = COPING/RAMP sign up questionnaire; !only of those participants who have reported a change in symptoms!

@@

+++CH: Create a column based on the the startDate if the participant can be counted as "prepandemic"
+++CH: Create three columns for all the definitions of "pre-pandemic" for sensitivy analysis

Potential dates for the definition of prepandemic
```{r Potential dates for prepandemic definition}
# Date of the first COVID-19 case in the United Kingdom (UK)
date_first_case_uk_january_31 <- as.POSIXct("2020-01-31")

@@

Gerome: We can get actual test data from PHE -> Cambridge
We need to move our data to Cambridge cluster
That's only for non-RAMP; EDGI, GLAD; NBR
KY: Longer term or for this paper?
Sanity check with Cambridge about the validity of the mysterious formula (Spector magic)

>>>>

This will is the preprocessing script to prepare the data to run analyses for the project "Anxiety, depression and trauma symptom change during the COVID-19 pandemic: retrospective versus objective assessment" - Young et al (2020)

@@

# Colour palettes
Define colours for plotting this are the standard coping colours
```{r Colour palettes: COPING}
COPINGpalette2 <- c("#78D9C5",
                    "#F5BE5E")

@@

# Data import and merging

Note: GLAD only receives current measures and RAMP questionnaire
Note: All COPING surveys are separate surveys

@@

#Clean Age variable
Age outlier
```{r age outlier}
# define age limits 
age_upper_limit = 117 # oldest person in the world is 117 years
age_lower_limit = 16

@@

```{r glad dem head}
head(glad.dem.raw.id)
```

```{r glad dem summary}
summarytools::dfSummary(
  glad.dem.raw.id,
  graph.col = F,
  valid.col = F,
  labels.col = F)
```

@@

# Grouping mental health disorders
*#Disorder groupings - from COPING/RAMP analysis minutes*
1. Depressive disorder group: major depression, antenatal/postnatal depression, premenstrual depression. The logic for this has been double checked for males - KT/MD. 
2. Anxiety group: generalised anxiety, social anxiety, specific phobia, agoraphobia, panic_disorder, panic_attacks. 
3. Agoraphobia and panic_disorder group: panic_disorder, agoraphobia, panic_attacks.
4. Eating disorders group: anorexianervosa, atypical anorexia nervosa, bulimia nervosa, overeating or bingeeating.
5. Obsessive complusive disorders: obsessive compulaive disorder, obsessive compulsive related disorders. 
6. Psychotic disorders: schizophrenia, schizoaffective disorder, general psychosis or psychotic illness. 
7. Personality disorders: cluster A, cluster B, cluster C
8. Control group: no diagnosed disorders.

@@

```{r glad mhq depressive_disorders}
glad.mhq.raw.id <- glad.mhq.raw.id %>%
  mutate(
    depressive_disorders_numeric =
      case_when(
        (mhq.depression_numeric == "0" &
        (mhq.pregnancy_depression_numeric == "0" | is.na(mhq.pregnancy_depression_numeric)) & 
        (mhq.premenstrual_dysphoric_disorder_pmdd_numeric == "0" | is.na(mhq.premenstrual_dysphoric_disorder_pmdd_numeric))) ~ 0,
        mhq.depression_numeric == "1" ~ 1,
        mhq.pregnancy_depression_numeric == "1" ~ 1,
        mhq.premenstrual_dysphoric_disorder_pmdd_numeric == "1" ~ 1))

@@

```{r Check agoraphobia_panic_disorder}
glad.mhq.raw.id %>%
  select(
    agoraphobia_panic_disorder,
    agoraphobia_panic_disorder_numeric,
    mhq.agoraphobia_numeric,
    mhq.panic_disorder_numeric,
    mhq.panic_attacks_numeric,
    mhq.agoraphobia,
    mhq.panic_disorder,
    mhq.panic_attacks
  )
```

@@

```{r Check psychotic_disorders}
glad.mhq.raw.id %>%
  select(
    psychotic_disorders,
    psychotic_disorders_numeric,
    mhq.schizophrenia_numeric,
    mhq.schizoaffective_disorder_numeric,
    mhq.type_psychosis_psychotic_illness_numeric,
    mhq.schizophrenia,
    mhq.schizoaffective_disorder,
    mhq.type_psychosis_psychotic_illness
  )
```

@@

@@

```{r pre, base, retro comparison for phq, pcl and gad}

```{r data level groups}

@@

Note: Retro for the PCL does not mean retrospective

```{r coping glad pcl cleaning}
coping.glad.pcl.raw.id <- coping.glad.pcl.raw %>%
  drop_na(externalDataReference) %>% # Drop NAs
  distinct(externalDataReference, .keep_all = TRUE) %>% # Remove duplicates based on ID
  separate(externalDataReference, into = c("Sample", "ID"), sep = 4) %>% # Split ID in Sample and Number
  mutate(ID = as.numeric(ID),
         startDate.glad = startDate) %>% # Rename startDate
  select(
    Sample, # Sample
    ID, # ID
    pcl.stressful_experience_repeated_images_base_numeric = pcl6.stressful_experience_repeated_images_numeric,
    pcl.stressful_experience_repeated_images_base = pcl6.stressful_experience_repeated_images,
    pcl.stressful_experience_upset_reminded_base_numeric = pcl6.stressful_experience_upset_reminded_numeric,
    pcl.stressful_experience_upset_reminded_base = pcl6.stressful_experience_upset_reminded,
    pcl.stressful_situation_avoiding_activities_base_numeric = pcl6.avoiding_activities_stressful_experience_numeric, # differ to GLAD prepan
    pcl.stressful_situation_avoiding_activities_base = pcl6.avoiding_activities_stressful_experience, # differ to GLAD prepan
    pcl.cut_people_feeling_distant_base_numeric = pcl6.cut_people_feeling_distant_numeric,
    pcl.cut_people_feeling_distant_base = pcl6.cut_people_feeling_distant,
    pcl.feeling_irritable_or_having_angry_outbursts_base_numeric = pcl6.feeling_irritable_or_having_angry_outbursts_numeric,
    pcl.feeling_irritable_or_having_angry_outbursts_base = pcl6.feeling_irritable_or_having_angry_outbursts,
    pcl.difficulty_concentrating_base_numeric = pcl6.difficulty_concentrating_numeric,
    pcl.difficulty_concentratin_base = pcl6.difficulty_concentrating,
    pcl.stressful_experience_repeated_images_retro_numeric = pcl6.stressful_experience_repeated_images.1_numeric,
    pcl.stressful_experience_repeated_images_retro = pcl6.stressful_experience_repeated_images.1,
    pcl.stressful_experience_upset_reminded_retro_numeric = pcl6.stressful_experience_upset_reminded.1_numeric,
    pcl.stressful_experience_upset_reminded_retro = pcl6.stressful_experience_upset_reminded.1,
    pcl.stressful_situation_avoiding_activities_retro_numeric = pcl6.avoiding_activities_stressful_experience.1_numeric,
    pcl.stressful_situation_avoiding_activities_retro = pcl6.avoiding_activities_stressful_experience.1,
    pcl.cut_people_feeling_distant_retro_numeric = pcl6.cut_people_feeling_distant.1_numeric,
    pcl.cut_people_feeling_distant_retro = pcl6.cut_people_feeling_distant.1,
    pcl.feeling_irritable_or_having_angry_outbursts_retro_numeric = pcl6.feeling_irritable_or_having_angry_outbursts.1_numeric,
    pcl.feeling_irritable_or_having_angry_outbursts_retro = pcl6.feeling_irritable_or_having_angry_outbursts.1,
    pcl.difficulty_concentrating_retro_numeric = pcl6.difficulty_concentrating.1_numeric,
    pcl.difficulty_concentratin_retro = pcl6.difficulty_concentrating.1
  ) %>%
  mutate_if(is.numeric, ~na_if(., -88)) %>% # Recode missing values to NAs in numeric variables
  mutate_if(is.numeric, ~na_if(., -99)) %>%
  mutate_if(is.numeric, ~na_if(., -77)) %>%
  mutate_if(is.factor, ~na_if(., "Seen but not answered")) %>% # Recode missing values to NAs in factor variables
  mutate_if(is.factor, ~na_if(., "Don't know")) %>%
  mutate_if(is.factor, ~na_if(., "Prefer not to say")) %>%
  mutate_if(is.factor, ~na_if(., "Prefer not to answer")) %>%
  mutate_if(is.factor, ~forcats::fct_drop(., only = "Seen but not answered")) %>% # Drop empty factor levels
  mutate_if(is.factor, ~forcats::fct_drop(., only = "Don't know")) %>%
  mutate_if(is.factor, ~forcats::fct_drop(., only = "Prefer not to answer")) %>%
  mutate_if(is.factor, ~forcats::fct_drop(., only = "Prefer not to say"))

@@

```{r coping glad phq summary labels}
summarytools::dfSummary(
  coping.glad.phq.raw,
  graph.col = F,
  valid.col = F,
  labels.col = F
)
```

@@

Exclusion of inflammatory bowel disease (IBD) BioResource (BR) cohort
```{r COVID exclusion of IBD}
#dim(dat.no.sex.no.age)
#
#dat.no.sex.no.age <- dat.no.sex.no.age %>%
#  filter(NBR.IBD == 0)
#
#dim(dat.no.sex.no.age.no.ibd)
#
#dim(dat.no.sex.no.age)[1]-dim(dat.no.sex.no.age.no.ibd)[1]

@@

#Missigness of questionnaire items
+++CH: We need to discuss how we want to exclude these participants that are missing answers to the questionnaires
```{r missingness}

```{r Exclusion of participants because of missing questionnaire items}
#dim(dat.no.sex.no.age.no.ibd)
#dat.no.sex.no.age.no.ibd.incomplete.questionnaire <- dat.no.sex.no.age.no.ibd %>%
#  filter(#ADD CONDITION HERE)
#dim(dat.no.sex.no.age.no.ibd.incomplete.questionnaire)
#
#dim(dat.no.sex.no.age.no.ibd)[1]-dim(dat.no.sex.no.age.no.ibd.incomplete.questionnaire)[1]

@@

# Definition of prepandemic and baseline 
Exlcude people who have joined GLAD/EDGI after the pandemic began, sensitivity analyses done on the cut off's we have chosen. 
Potentially GLAD/EDGI particiapnts between January to April. EDGI launched on February 24th.

@@

Retrospective = COPING/RAMP sign up questionnaire; *only of those participants who have reported a change in symptoms*

@@

Dates for the definition of prepandemic
```{r Potential dates for prepandemic definition}
# Date of the first COVID-19 case in the United Kingdom (UK)
date_first_case_uk_january_31 <- as.POSIXct("2020-01-31")

@@
